\chapter{Acknowledgements}

\textit{Where to begin?}
\\
Thank you to my supervisers, Ed Green and Sanjeeva Balasuriya,
for your enduring support during these \textit{two and a half} 
years.
\\
Thankyou to my highschool English teacher, Tanja, for teaching
me how to use form and technique in my writing.
\\
Thank you to my parents, Debbie and Alex, and to my sister,
Amelia. In your own way, you each provided the soil 
from which this thesis grew.
\\
Thank you to my literary friend, John, whose cerebral writing 
will have inspired a generation of \textit{rebels with a cause} by 
next Tuesday, I'm sure of it.
\\
Thank you to Ki and my other colleagues at \textit{Adelaide University}
for providing the environmental stresses to keep me growing. Ki 
was always willing to speak with me and offer his time. 
I am grateful.
\\
There appears to be a collection of \textit{Alex}'s in my life.
I thank you all dearly in one foul sweep. 
\\
To \textit{Radee}, for reminding be how to thinking critically 
and creatively at the same time.
\\
To Harry and Zehao and the other engineers one of whom is another \textit{Alex},
I appreciate you.
\\
To Giuseppe, thank you for introducing me to the \textit{bible of statistical mechanics}.
\\
For the fruitful \textit{Kong-Tchorbadjiev-Nakone} (KTN) correspondence,
for your years of philosphical musing.
\\
Thank you to \textit{cafe} staff at the university for your 
unassuming presence among the day-to-day emergence of new forms.
\\
Thankyou to the late philosophy Professor Michael Sugrue,
whose lectures, both young and old, were stunning depictions
of daring heart of philosophy.
\\
Thankyou to \textit{ChatGPT},
for helping me externalise my interior geometry.
It materialised in the form of the great parable of baker's yeast.
\\
If there was anyone I forgot, I hope you are not offended.
\\

\textit{Isaac}
